\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis

### 2023-01-18 src/tex/ms.tex
Adding citations to intro

% Abstract with filler text
\begin{abstract}
    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    As more detections continue, data analysis taks are becoming increasingly computationally demanding. \kw{I would cut everything before out.}
    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to tradiaitional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughtout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.

@@

% Intro outline
% - What are gravitational waves and what are data analysis tasks?
% - What are derivatives, and how can they be used?
% - What is automatic differentiation, and why has it become particularly useful now? JAX?
% - Discussion of what waveforms are most ameanable to being differentiable
% - In this paper

The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
In addition, this computational cost will substatially increase when next generation detectors come online.

The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event.
In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event. \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity.
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{Wong:2022flg}. \kw{Need more reference and some rewording}
All of these tasks require significant compute. \kw{This could be triggering to some people, since population analysis could be quite cheap}
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.

Derivatives are ubiqitously useful throughout data analysis tasks.
For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g. in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain. \kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calcuations for gravitational waveforms~\citep{Coogan:2022qxs}.

>>>>

% Abstract with filler text
\begin{abstract}
    Since their discovery in 2015, gravitaitonal waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    As more detections continue, data analysis taks are becoming increasingly computationally demanding.
    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analsyis tasks.
    Firstly, we demonstrate that highly accurate waveforms (IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that make up the waveform. 
    In particular, we demonstrate that the typical \textit{match} with numerical relativiely waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to tradiaitional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughtout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.

@@

The discovery of gravitational waves (GWs) from inspiraling and mergering compact objects (COs) has revolutionized our understanding of both fundamental physics and astronomy~\citep{LIGOScientific:2021djp}.
Although the data volumes from GW detectors are relatively small, analyzing the data is a computationally demanding task.
In addition, this computational cost will substatially increase when next generation detectors come online~\citep{Maggiore:2019uih, Reitze:2019iox}.

The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. 
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
In the most minimial example, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
Beyond this simple scenario, more complex waveform models with many additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio}.
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, Wong:2022flg}.
All of these tasks require significant compute.
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleaviating this computational demand.

Derivatives are ubiqitously useful throughout data analysis tasks.
For instance, during PE, derivative information can be used to guide the sampling towards regions with higher likelihood values (e.g. in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Decsent).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain.
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calcuations for gravitational waveforms~\citep{Coogan:2022qxs}.
